### 2023-10-27 records/manuscript_2023.tex
Improve Figs. 4-5, polish code

% evidence for the model (PE neurons, connectivity PE --> M, predictions or P neurons exist, P neurons may hold the mean, time constants chnage along hierarchy)

@@

@@

@@

XXX Please note that xxx model parameters chosen do not change overall results are not particular to get the results xxx. Important factors are: 1) long enough to get to steady state (as PE circuits used were learned for steady state, 2) lower M updates faster than higher M, 3) update speed for M and V (see lambda and theta etc.) must be chosen in accordance with trial duration stimulus duration. XXX This could also go to the section Simulations xxx

@@

In the following, we give all figure-specific parameters not directly visible or mentioned in the Figures and captions. Please note that the qualitative results are robust to the choice of the simulation parameters and are here stated merely to ensure the reproducibility of all figures. Furthermore, to increase readability, we do not include units for the parameters. All units can be deduced from the equations above.\newline\\ 

@@

\textbf{Figure 1:} Simulation time steps = $10^5$, constant prediction (fixed) = $5$, input mean $\in [0,10]$, input standard deviation $0$.\\

@@

\textbf{Figure 2:} Simulation time steps = $10^5$, $200$ constant values (each $500$ time steps long) drawn from a uniform distribution, B and C: input standard deviation fixed at $4.5$ when mean is varied, input mean fixed at $4.5$ when input variance is varied.\\

@@

\textbf{Figure 2 Supplementary Fig. 1:} Simulation time steps = $10^5$, $200$ constant values (each $500$ time steps long) drawn from a uniform distribution with mean of $5$ and variance of $4$. Number of repetitions with different seeds: 20.\\

@@

\textbf{Figure 2 Supplementary Fig. 2:} Number of time steps were $10^5$ with $200$ constant values drawn from a uniform distribution with mean of $5$ and variance of $4$. Time step was $0.1$. The connections from the PE neurons to the M or V neuron were altered by a factor $\gamma$ drawn from a normal distribution. If not stated otherwise (in the Figure), the mean of this normal distribution was $1$ and the variance $0$, while the connection probability was $1$. Number of repetitions with different seeds: 10.\\

@@

\textbf{Figure 2 Supplementary Fig. 3:} Number of time steps were $4000$ (XXX maybe mention here and other above stating "except otherwise stated" or so XXX). Number of identical mean-field networks: $1000$.  Simulation time steps = $10^5$.\\

@@

\textbf{Figure 3:} $100$ trials, each simulated for $5000$ time steps. In a trail, $10$ constant values were drawn from a normal distribution (each $500$ time steps long). The variance of this distribution was $0$ (C, F)  and $5$ (D). In each trial, the stimulus mean was drawn from an uniform distribution $U(a, b)$ (C: $U(1,9)$, D: $U(5,5)$). In F, $U(a,b)$ was parameterized such that the trail variability was $3$.\\

@@

\textbf{Figure 3 Supplementary Fig. 1:}  $100$ trials, each simulated for $5000$ time steps. In a trail, $10$ constant values were drawn from a normal distribution $N(\mu_\mathrm{in}, \sigma_\mathrm{in})$, each $500$ time steps long. In each trial, the stimulus mean was drawn from an uniform distribution $U(a, b)$. Switch of input statistics occurs after 50 trails. State 1: stimulus $\in N(\mu_\mathrm{in}, 0)$ with $\mu_\mathrm{in} \in U(5,5)$, State 2:  stimulus $\in N(\mu_\mathrm{in}, 3)$ with $\mu_\mathrm{in} \in U(5,5)$, State 3: stimulus $\in N(\mu_\mathrm{in}, 3)$ with $\mu_\mathrm{in} \in U(0,10)$, State 4: stimulus $\in N(\mu_\mathrm{in}, 0)$ with $\mu_\mathrm{in} \in U(0,10)$.\\

@@

\textbf{Figure 3 Supplementary Fig. 2:} $200$ trials, each simulated for $5000$ time steps. In a trail, $10$ constant values were drawn from a normal distribution with variance $\sigma^2_\mathrm{in}$, each $500$ time steps long. In each trial, the stimulus mean was drawn from an uniform distribution with mean of $5$ and a variance of $\sigma^2_\mathrm{trial}$. $\sigma^2_\mathrm{trial} \in \lbrace0, 0.75, 1.5, 2.25, 3\rbrace$, $\sigma^2_\mathrm{in} \in \lbrace 3, 2.25, 1.5,0.75, 0\rbrace$. A: scaling factors of $w_\mathrm{M\leftarrow PE}$ were 0.3 and 6.7 XXX.\\

@@

\textbf{Figure 4:} $100-200$ trials, each simulated for $5000$ time steps. In a trail, $10$ constant values were drawn from a normal distribution with variance $\sigma^2_\mathrm{in}$, each $500$ time steps long. In each trial, the stimulus mean was drawn from an uniform distribution with mean of $5$ and a variance of $\sigma^2_\mathrm{trial}$. A, top: $\sigma^2_\mathrm{in} = 0$ and $\sigma^2_\mathrm{trial} = 1$. A, bottom: $\sigma^2_\mathrm{in} = 1$ and $\sigma^2_\mathrm{trial} = 0$. C: $\sigma^2_\mathrm{in} = 1$ and $\sigma^2_\mathrm{trial} = 1$. D: $\sigma^2_\mathrm{in} = 5.3 XXX$ and $\sigma^2_\mathrm{trial} = 0$. Additional input (perturbation) was either fixed at $1$ (A, D) or systematically varied between $-1$ and $1$ (C). To estimate changes in baseline and gain of nPE and pPE neurons, we fitted a linear function to the PE neuron activity for the input range $[0, 2.5]$.\\

@@

\textbf{Figure 5:} $200$ trials, each simulated for $5000$ time steps (except in panel E). In a trail, $10$ constant values were drawn from a normal distribution (each $500$ time steps long) with standard deviation $\sigma_\mathrm{in}$. In each trial, the stimulus mean $\mu_\mathrm{in}$ was drawn from an uniform distribution $U(a, b)$. A: $\sigma_\mathrm{in} \in \lbrace1,7\rbrace$, $\mu_\mathrm{in} \in U(15, 25)$. B: $\sigma_\mathrm{in} \in [0,8]$, $\mu_\mathrm{in} \in U(15, 25)$, and $\sigma_\mathrm{in} = 5$, $\mu_\mathrm{in} \in U(15, b)$ with $b\in [20, 48]$. C: $\sigma_\mathrm{in} \in \lbrace2,5\rbrace$, $\mu_\mathrm{in} \in U(15, 15)$. D: $\sigma_\mathrm{in} = 0$, $\mu_\mathrm{in} \in U(15, 25)$ or $U(10, 30)$. E: $\sigma_\mathrm{in} = 0$, $\mu_\mathrm{in} \in U(15, 25)$, or $\sigma_\mathrm{in} = 5$, $\mu_\mathrm{in} \in U(15, 15)$,  Time steps per trail increased from $5000$ to $10^4$.\\

@@

\textbf{Figure 5 Supplementary Fig. 1:} $200$ trials, each simulated for $5000$ time steps. In a trail, $10$ constant values were drawn from a normal distribution (each $500$ time steps long) with standard deviation $\sigma_\mathrm{in}$. In each trial, the stimulus mean $\mu_\mathrm{in}$ was drawn from an uniform distribution $U(a,b)$. We used two different uniform distributions $U(15, 25)$ and  $U(25, 35)$, and introduced scalar variability so that $\sigma_\mathrm{in}$ is a linear function of $\mu_\mathrm{in}$. Specifically, we chose $\sigma_\mathrm{in} = \mu_\mathrm{in} -14$.

@@

\subsection{Comparison to Kalman filter and Bayes Factor surprise}
%
Kalman filter. Initialisation
%
\begin{align*}
x_{0|init} &= 0 \\
P_{0|init} &= \sigma^2\ I
\end{align*}
%
with x being the system state (in my terms the prediction), P is the covariance matrix of the errors of x (in my terms the var of the predictions) and I is the identity matrix.
%
Then the "correction" is given by
%
\begin{align*}
K_k &= P_{k|k-1}\ H_k^T\ \left( H_k\ P_{k|k-1}\ H_k^T + R_k \right)^{-1} \\
x_k &= x_{k|k-1} + K_k\ \left( z_k - H_k\ x_{k|k-1}\right) \\
P_k &= \left( I - K_k\ H_k\right)\ P_{k|k-1}
\end{align*}
%
with K the kalman gain matrix, H the observation matrix ($z_k = H_k\ x_k + noise$), R the covaraince of the measurement noise and z a new observation. The last part of the Kalman filter is the "prediction":
%
\begin{align*}
x_{k|k-1} &= F_{k-1}\ x_{k-1} + B_{k-1}\ u_{k-1} \\
P_{k|k-1} &= F_{k-1}\ P_{k-1}\ F_{k-1}^T + Q_{k-1}
\end{align*}
%
with F the transition matrix ($x_{k|k-1} = F_{k-1}\ x_{k-1}$, u a deterministic perturbation, B the dynamics of the deterministic perturbation. In our terms
%
\begin{align*}
\alpha = K_k = \frac{P_{k|k-1}}{R_k + P_{k|k-1}}
\end{align*}
%
$P_{k|k-1}$, is however $\sigma_P^2$ in my implementation and $R_k$ is fixed variance of inputs $\sigma_S^2$. Hence, my implementation represents (?) the Kalman filter. Important to note is, that in my implementation we estimate the variance of inputs dynamically, so it is not set! Another nice advantage here is that I don't need a good estimate for P. I can basically initiate it as I want. Another difference is that I consider the optimal weighting in my "output neuron" and not the prediction itself ... .

XXX Comparison to Bayes Factor surprise

>>>>

% evidence for the model (PE neurons, connectivity PE --> M, predictions or P neurons exist, P neurons may hold the mean, time constants change along hierarchy)

@@

% selectivity? One-dimensional system (wahrscheinlich nicht)

@@

----
% Katha feedback: predictions, postdictions, experiments vorschlagen, limitations, grosserer kontext, abgrenzung zu reward prediction error

@@

@@

The qualitative results were fairly robust to the choice of the simulation parameters and are here stated merely to ensure the reproducibility of all figures. However, we note that we made use of PE circuits that had been trained on steady state inputs \ref{XXX}. Hence, we must simulate the network long enough to ensure that the PE neurons reach their steady state. Moreover, the lower-level M neuron must evolve faster than the higher-level M neuron as indicated in Fig. \ref{XXX}. Finally, the time constant of the V neurons must be of the same magnitude as the trial duration.

In the following, we give all figure-specific parameters not directly visible or mentioned in the figures and captions. Furthermore, to increase readability, we do not include units for the parameters. All units can be deduced from the equations above. We simulated the network in Figures \ref{fig:Fig_1} and \ref{fig:Fig_2} for $10^5$ simulation time steps. In Figure \ref{fig:Fig_2}, we presented $200$ constant values, each $500$ time steps long. In Figure \ref{fig:Fig_3} (including supporting figures) we simulated $100$ trials, while in Figures \ref{fig:Fig_4} to \ref{fig:Fig_5} (including supporting figures), we simulated $200$ trials, each $5000$ time steps long. In a trail, $10$ constant values were drawn from a normal distribution $N(\mu_\mathrm{in}, \sigma^2_\mathrm{in})$, each $500$ time steps long. The stimulus mean was drawn from an uniform distribution, $U(a, b)$, with mean $\mu_\mathrm{in}$ and variance $\sigma^2_\mathrm{trial}$.\newline\\ 

@@

\textbf{Figure 1:} Constant prediction (fixed) = $5$, input mean $\in [0,10]$, input standard deviation $0$.\\

@@

\textbf{Figure 2:} Inputs drawn from a uniform distribution. B and C: input standard deviation fixed at $4.5$ when mean is varied, input mean fixed at $4.5$ when input variance is varied.\\

@@

\textbf{Figure 2 Supplementary Fig. 1:} Inputs drawn from different distributions with mean of $5$ and variance of $4$. Number of repetitions with different seeds: 20.\\

@@

\textbf{Figure 2 Supplementary Fig. 2:} Inputs drawn from a uniform distribution with mean of $5$ and variance of $4$. Time step was $0.1$. The connections from the PE neurons to the M or V neuron were altered by a factor $\gamma$ drawn from a normal distribution. If not stated otherwise, the mean of this normal distribution was $1$ and the variance $0$, while the connection probability was $1$. Number of repetitions with different seeds: 10.\\

@@

\textbf{Figure 2 Supplementary Fig. 3:} Number of time steps were $4000$. Number of identical mean-field networks: $1000$. \\

@@

\textbf{Figure 3:} C: $\sigma^2_\mathrm{in} = 0$ and $U(1,9)$, D:  $\sigma^2_\mathrm{in} = 5$ and $U(5,5)$, F: $\sigma^2_\mathrm{in} = 0$ and $U(a,b)$ was parameterized such that the trail variability was $3$.\\

@@

\textbf{Figure 3 Supplementary Fig. 1:}  Switch of input statistics occurs after 50 trails. State 1: stimulus $\in N(\mu_\mathrm{in}, 0)$ with $\mu_\mathrm{in} \in U(5,5)$, State 2:  stimulus $\in N(\mu_\mathrm{in}, 3)$ with $\mu_\mathrm{in} \in U(5,5)$, State 3: stimulus $\in N(\mu_\mathrm{in}, 3)$ with $\mu_\mathrm{in} \in U(0,10)$, State 4: stimulus $\in N(\mu_\mathrm{in}, 0)$ with $\mu_\mathrm{in} \in U(0,10)$.\\

@@

\textbf{Figure 3 Supplementary Fig. 2:} $\mu_\mathrm{in} = 5$ and $\sigma^2_\mathrm{trial} \in \lbrace0, 0.75, 1.5, 2.25, 3\rbrace$, $\sigma^2_\mathrm{in} \in \lbrace 3, 2.25, 1.5,0.75, 0\rbrace$. A: scaling factors of $w_\mathrm{M\leftarrow PE}$ were 0.3 and 7.\\

@@

\textbf{Figure 4:} $\mu_\mathrm{in} = 5$. A, top: $\sigma^2_\mathrm{in} = 0$ and $\sigma^2_\mathrm{trial} = 1$. A, bottom: $\sigma^2_\mathrm{in} = 1$ and $\sigma^2_\mathrm{trial} = 0$. C: $\sigma^2_\mathrm{in} = 1$ and $\sigma^2_\mathrm{trial} = 1$. D: $\sigma^2_\mathrm{in} = 5$ and $\sigma^2_\mathrm{trial} = 0$. Additional input (perturbation) was either fixed at $0.5$ (A, D) or systematically varied between $-1$ and $1$ (C), and was \textit{on} for the last 50\% of the trials. To estimate changes in baseline and gain of nPE and pPE neurons, we fitted a linear function to the PE neuron activity for the input range $[0, 2.5]$.\\

@@

\textbf{Figure 5:} A: $\sigma_\mathrm{in} \in \lbrace1,7\rbrace$, $\mu_\mathrm{in} \in U(15, 25)$. B: $\sigma_\mathrm{in} \in [0,8]$, $\mu_\mathrm{in} \in U(15, 25)$, and $\sigma_\mathrm{in} = 5$, $\mu_\mathrm{in} \in U(15, b)$ with $b\in [20, 48]$. C: $\sigma_\mathrm{in} \in \lbrace2,5\rbrace$, $\mu_\mathrm{in} \in U(15, 15)$. D: $\sigma_\mathrm{in} = 0$, $\mu_\mathrm{in} \in U(15, 25)$ or $U(10, 30)$. E: $\sigma_\mathrm{in} = 0$, $\mu_\mathrm{in} \in U(15, 25)$, or $\sigma_\mathrm{in} = 5$, $\mu_\mathrm{in} \in U(15, 15)$,  Time steps per trail increased from $5000$ to $10^4$.\\

@@

\textbf{Figure 5 Supplementary Fig. 1:} We used two different uniform distributions $U(15, 25)$ and  $U(25, 35)$, and introduced scalar variability so that $\sigma_\mathrm{in}$ is a linear function of $\mu_\mathrm{in}$. Specifically, we chose $\sigma_\mathrm{in} = \left[\mu_\mathrm{in} -14\right]_+$.

@@

%\subsection{Comparison to Kalman filter and Bayes Factor surprise}
%%
%Kalman filter. Initialisation
%%
%\begin{align*}
%x_{0|init} &= 0 \\
%P_{0|init} &= \sigma^2\ I
%\end{align*}
%%
%with x being the system state (in my terms the prediction), P is the covariance matrix of the errors of x (in my terms the var of the predictions) and I is the identity matrix.
%%
%Then the "correction" is given by
%%
%\begin{align*}
%K_k &= P_{k|k-1}\ H_k^T\ \left( H_k\ P_{k|k-1}\ H_k^T + R_k \right)^{-1} \\
%x_k &= x_{k|k-1} + K_k\ \left( z_k - H_k\ x_{k|k-1}\right) \\
%P_k &= \left( I - K_k\ H_k\right)\ P_{k|k-1}
%\end{align*}
%%
%with K the kalman gain matrix, H the observation matrix ($z_k = H_k\ x_k + noise$), R the covaraince of the measurement noise and z a new observation. The last part of the Kalman filter is the "prediction":
%%
%\begin{align*}
%x_{k|k-1} &= F_{k-1}\ x_{k-1} + B_{k-1}\ u_{k-1} \\
%P_{k|k-1} &= F_{k-1}\ P_{k-1}\ F_{k-1}^T + Q_{k-1}
%\end{align*}
%%
%with F the transition matrix ($x_{k|k-1} = F_{k-1}\ x_{k-1}$, u a deterministic perturbation, B the dynamics of the deterministic perturbation. In our terms
%%
%\begin{align*}
%\alpha = K_k = \frac{P_{k|k-1}}{R_k + P_{k|k-1}}
%\end{align*}
%%
%$P_{k|k-1}$, is however $\sigma_P^2$ in my implementation and $R_k$ is fixed variance of inputs $\sigma_S^2$. Hence, my implementation represents (?) the Kalman filter. Important to note is, that in my implementation we estimate the variance of inputs dynamically, so it is not set! Another nice advantage here is that I don't need a good estimate for P. I can basically initiate it as I want. Another difference is that I consider the optimal weighting in my "output neuron" and not the prediction itself ... .
%
%XXX Comparison to Bayes Factor surprise
